Assessing primitives performance\\ on multi-stage execution

### 2018-05-10 paper.tex
added comments

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} 
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT and allows the benchmarks to run at least 1 second in the slower implementations.
The loop overhead is measured and removed from the measurements.

The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations \footnote{There is one branch less in the SlangPlugin implementation in this case}.

\subsection{Micro-benchmarks}

@@

>>>>

\cba{We may use paragraph instead of itemize here to compact}
 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (-) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} 
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor noises}.
The loop overhead is measured and removed from the measurements.

The 2 compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.
I
\subsection{Micro-benchmarks}

@@

\cba{We need to explain: small string, overhead stack switch high, baseline and JIT version fastest. Large string, loop is where the time is spent, slang and JIT fastest.
We need size 1000 so that the stack switch overhead cannot be seen anymore on gcc (size 100 we still C it)
Size 0 is only '' = '', not really interesting in real app, but show well the stack switch overhead.
Example use cases are comparing tokens in parsers (Size ~4-10 [self, super, thisContext, this, etc.])
Other example are long urls, hundreds of characters, which are represented with large string comp bench}
